---
title: "A negative binomial renewal model"
output: html_notebook
---

```{r}
library(incidenceinflation)
library(ggplot2)
library(magrittr)
library(dplyr)
library(tidyr)
library(purrr)
```


In this vignette, we show how to fit our model to data assuming a negative binomial renewal model.

```{r}
days_total <- 100

# allow Rt to vary over time and construct interpolation function
v_Rt <- c(rep(1.5, 40), rep(0.4, 20), rep(1.5, 40))
Rt_function <- stats::approxfun(1:days_total, v_Rt)

# serial interval distribution parameters (assumed gamma)
s_params <- list(mean=5, sd=3)

# negative binomial over-dispersion parameter
kappa <- 8

# specify a reporting distribution that shortens after the first 50 days
mean_1 <- 10
mean_2 <- 3
r_params <- tibble(
  time_onset=seq(1, days_total, 1),
  mean=c(rep(mean_1, 50), rep(mean_2, 50)),
  sd=c(rep(3, 50), rep(2, 50))
)

# generate data
df <- generate_snapshots(days_total, Rt_function, s_params, r_params,
                         kappa=kappa)
```

With a negative binomial model, the cases evolve more stochastically.
```{r}
df_true <- df %>% 
  group_by(time_onset) %>% 
  summarise(cases_true=last(cases_true))
df_true %>% 
  ggplot(aes(x=time_onset, y=cases_true)) +
  geom_line()
```

Performing inference assuming an (incorrect) Poisson model.
```{r}
snapshot_with_Rt_index_df <- df
initial_cases_true <- df %>% select(time_onset, cases_true) %>% unique()

reporting_pieces <- tibble(
  time_onset=unique(df$time_onset),
  reporting_piece_index=c(rep(1, 50), rep(2, 50))
)
snapshot_with_Rt_index_df <- snapshot_with_Rt_index_df %>%
  select(-cases_true) %>% 
  left_join(reporting_pieces, by="time_onset")
initial_Rt <- tribble(~Rt_index, ~Rt,
                      1, 1.5,
                      2, 1.5,
                      3, 0.4,
                      4, 1.5,
                      5, 1.5)
Rt_indices <- unlist(map(seq(1, 5, 1), ~rep(., 20)))

Rt_index_lookup <- tibble(
  time_onset=seq_along(Rt_indices),
  Rt_index=Rt_indices)
snapshot_with_Rt_index_df <- snapshot_with_Rt_index_df %>% 
  left_join(Rt_index_lookup)

initial_reporting_parameters <- tibble(
  reporting_piece_index=unique(reporting_pieces$reporting_piece_index)
) %>% 
  mutate(
    mean=c(5, 5),
    sd=3
  )
serial_parameters <- list(mean=5, sd=3)
Rt_prior <- list(shape=5, rate=5)
priors <- list(Rt=Rt_prior,
               reporting=list(mean_mu=5,
                              mean_sigma=10,
                              sd_mu=3,
                              sd_sigma=5),
               max_cases=10000)

res <- mcmc(niterations=100,
            snapshot_with_Rt_index_df,
            priors,
            serial_parameters,
            initial_cases_true,
            initial_reporting_parameters,
            initial_Rt,
            reporting_metropolis_parameters=list(mean_step=0.25, sd_step=0.1),
            serial_max=40, p_gamma_cutoff=0.99, maximise=FALSE,
            nchains = 1, is_parallel = FALSE)
```

Comparing estimated cases with true, we find a poor approximation.
```{r}
cases_df <- res$cases
cases_sum <- cases_df %>% 
  group_by(time_onset) %>% 
  summarise(
    lower=quantile(cases_true, 0.025),
    middle=quantile(cases_true, 0.5),
    upper=quantile(cases_true, 0.975)
) %>% 
  mutate(cases_true=df_true$cases_true)
cases_sum %>% 
  filter(time_onset >= 75) %>% 
  ggplot(aes(x=time_onset)) +
  geom_ribbon(aes(ymin=lower, ymax=upper),
              fill="blue", alpha=0.4) +
  geom_line(aes(y=middle), colour="blue") +
  geom_line(aes(y=cases_true))
```
Now fitting a model assuming negative binomial renewal model.
```{r}
priors$overdispersion <- list(mean=5, sd=10)
res_1 <- mcmc(niterations=100,
            snapshot_with_Rt_index_df,
            priors,
            serial_parameters,
            initial_cases_true,
            initial_reporting_parameters,
            initial_Rt,
            reporting_metropolis_parameters=list(mean_step=0.25, sd_step=0.1),
            serial_max=40, p_gamma_cutoff=0.99, maximise=FALSE,
            nchains = 1, is_parallel = FALSE,
            is_negative_binomial = TRUE,
            overdispersion_metropolis_sd = 5)
```

Examine approach to overdispersion parameter
```{r}
res_1$overdispersion %>% 
  ggplot(aes(x=iteration, y=overdispersion)) +
  geom_line()
```


```{r}
cases_df <- res_1$cases
cases_sum_nb <- cases_df %>% 
  group_by(time_onset) %>% 
  summarise(
    lower=quantile(cases_true, 0.025),
    middle=quantile(cases_true, 0.5),
    upper=quantile(cases_true, 0.975)
) %>% 
  mutate(cases_true=df_true$cases_true) %>% 
  mutate(model="negative binomial")
cases_sum <- cases_sum %>% 
  mutate(model="Poisson")

cases_sum %>% 
  bind_rows(cases_sum_nb) %>% 
  filter(time_onset >= 75) %>% 
  ggplot(aes(x=time_onset)) +
  geom_ribbon(aes(ymin=lower, ymax=upper),
              fill="blue", alpha=0.4) +
  geom_line(aes(y=middle), colour="blue") +
  geom_line(aes(y=cases_true)) +
  facet_wrap(~model)
```

Examine approach to reporting parameters by Markov chain.
```{r}
res_1$reporting %>% 
  ggplot(aes(x=iteration, y=mean)) +
  geom_line() +
  geom_hline(data=tibble(means=c(mean_1, mean_2),
                         reporting_piece_index=c(1, 2)),
                         aes(yintercept=means),
             linetype=2) +
  facet_wrap(~reporting_piece_index)
```
```{r}
rt_nb <- res_1$Rt %>% 
  group_by(Rt_index) %>% 
  summarise(
    sd=sd(Rt),
    Rt=median(Rt)) %>% 
  mutate(model="negative binomial")
rt_poisson <- res$Rt %>% 
  group_by(Rt_index) %>% 
  summarise(
    sd=sd(Rt),
    Rt=median(Rt)) %>% 
  mutate(model="Poisson")
rt_nb
rt_poisson
```

