---
title: "Simulating data and fitting a model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{simulating_data_and_model_fit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(incidenceinflation)
library(ggplot2)
library(magrittr)
library(dplyr)
library(tidyr)
```

We first generate a true case series using a negative binomial renewal process.
Here, we assume a time-varying Rt.
```{r}
days_total <- 100

# allow Rt to vary over time and construct interpolation function
v_Rt <- c(rep(1.5, 40), rep(0.4, 20), rep(1.5, 40))
Rt_function <- stats::approxfun(1:days_total, v_Rt)

# serial interval distribution parameters (assumed gamma)
s_params <- list(mean=5, sd=3)

# negative binomial over-dispersion parameter
kappa <- 10

# generate series
cases <- true_cases(days_total, Rt_function, kappa, s_params,
                    initial_parameters=list(mean=30, length=30))

# plot series
tibble(d=1:days_total,
       cases=cases,
       Rt=v_Rt) %>%
  ggplot(aes(x=d)) +
  geom_point(aes(y=cases)) +
  geom_line(aes(y=cases)) +
  geom_line(aes(y=Rt * 500), colour="orange") +
  scale_y_continuous(sec.axis = sec_axis(~./500, name = "Rt"))
```
In our model, we assume that cases arising on a given day aren't necessarily observed until some time in the future. In other words, there are imperfections in case reporting meaning that cases may not be uncovered until much after they arise.

We now use the true case series generated previous to simulate observed case trajectories: there is a trajectory for each onset day.
```{r}
# reporting delay distribution (assumed gamma)
reporting_delay <- list(mean=10, sd=3)

df <- observed_cases(cases, reporting_delay, days_max_follow_up=40)
glimpse(df)
```

For example, consider those cases arising on day 10. We can plot the number of reported cases which arised on that day subsequently. Over time, reported cases tends to true case counts as move cases arising on day 10 are retroactively uncovered.
```{r}
df %>% 
  filter(time_onset == 10) %>% 
  pivot_longer(cols=c("cases_reported", "cases_true")) %>% 
  ggplot(aes(x=time_reported, y=value, colour=name)) +
  geom_line() +
  scale_color_brewer("Cases", palette = "Dark2")
```
Now plotting the same but for each onset day, we obtain a series of trajectories for reported cases.
```{r}
df %>% 
  ggplot(aes(x=time_reported, y=cases_reported,
             group=as.factor(time_onset))) +
  geom_line()
```

Another way of understanding these data is that the historical snapshots of the epidemic change as time goes on and more cases are retroactively reported. We imagine we are at $t=40$ and consider how the history of cases changes as time goes on.
```{r}
t <- 40
df %>% 
  filter(time_onset <= t) %>% 
  filter(time_reported >= t) %>% 
  ggplot(aes(x=time_onset, y=cases_reported)) +
  geom_line(aes(colour=time_reported,
                group=time_reported)) +
  geom_point(aes(y=cases_true), colour="black") +
  scale_colour_viridis_c(begin = 0.2)
```

