---
title: "simulating_data_and_model_fit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{simulating_data_and_model_fit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(incidenceinflation)
library(ggplot2)
library(magrittr)
library(dplyr)
```

We first generate a true case series using a negative binomial renewal process.
Here, we assume a time-varying Rt
```{r}
days_total <- 100

# allow Rt to vary over time and construct interpolation function
v_Rt <- c(rep(1.5, 40), rep(0.4, 20), rep(1.5, 40))
Rt_function <- stats::approxfun(1:days_total, v_Rt)

# serial interval distribution parameters (assumed gamma)
s_params <- list(mean=5, sd=3)

# negative binomial over-dispersion parameter
kappa <- 2

# generate series
cases <- true_cases(days_total, Rt_function, kappa, s_params)

# plot series
tibble(d=1:days_total,
       cases=cases,
       Rt=v_Rt) %>%
  ggplot(aes(x=d)) +
  geom_point(aes(y=cases)) +
  geom_line(aes(y=cases)) +
  geom_line(aes(y=Rt * 10), colour="orange") +
  scale_y_continuous(sec.axis = sec_axis(~./10, name = "Rt"))
```
In our model, we assume that cases arising on a given day aren't necessarily observed until some time in the future: then we suppose those cases are summised to have originated on the correct day. In other words, there are imperfections in case reporting meaning that cases may not be uncovered until much after they arise.

We now use the true case series generated previous to simulate observed case trajectories: there is a trajectory for each onset day.
```{r}
# reporting delay distribution (assumed gamma)
reporting_delay <- list(mean=10, sd=3)

observed_cases(cases, reporting_delay) %>% 
  ggplot(aes(x=time_reporting, y=cases_reported,
             group=as.factor(time_onset))) +
  geom_line()
```

