---
title: "Temporal variation in reporting delays"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{temporal_variation_in_reporting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(incidenceinflation)
library(ggplot2)
library(magrittr)
library(dplyr)
library(tidyr)
library(purrr)
```

In a real epidemic, the reporting delays may vary over time. In this vignette, we illustrate how this package can be used to simulate then fit a model allowing for temporally varying delays in case reporting.

```{r}
days_total <- 100

# allow Rt to vary over time and construct interpolation function
v_Rt <- c(rep(1.5, 40), rep(0.4, 20), rep(1.5, 40))
Rt_function <- stats::approxfun(1:days_total, v_Rt)

# serial interval distribution parameters (assumed gamma)
s_params <- list(mean=5, sd=3)

# negative binomial over-dispersion parameter
kappa <- 10

# specify a reporting distribution that shortens after the first 50 days
mean_1 <- 10
mean_2 <- 3
r_params <- tibble(
  time_onset=seq(1, days_total, 1),
  mean=c(rep(mean_1, 50), rep(mean_2, 50)),
  sd=c(rep(3, 50), rep(2, 50))
)

# generate data
df <- generate_snapshots(days_total, Rt_function, s_params, r_params,
                         kappa=kappa)
```

At t=40, there is large discrepancies between the cases reported and the true cases.
```{r}
t <- 40
df %>% 
  filter(time_onset <= t) %>% 
  filter(time_reported >= t) %>% 
  ggplot(aes(x=time_onset, y=cases_reported)) +
  geom_line(aes(colour=time_reported,
                group=time_reported)) +
  geom_point(aes(y=cases_true), colour="black") +
  geom_line(aes(y=cases_true), colour="black") +
  scale_colour_viridis_c(begin = 0.2)
```

At t=90, these are much less marked.
```{r}
t <- 90
df %>% 
  filter(time_onset <= t) %>% 
  filter(time_reported >= t) %>% 
  ggplot(aes(x=time_onset, y=cases_reported)) +
  geom_line(aes(colour=time_reported,
                group=time_reported)) +
  geom_point(aes(y=cases_true), colour="black") +
  geom_line(aes(y=cases_true), colour="black") +
  scale_colour_viridis_c(begin = 0.2)
```

Performing inference
```{r}
snapshot_with_Rt_index_df <- df
initial_cases_true <- df %>% select(time_onset, cases_true) %>% unique()

reporting_pieces <- tibble(
  time_onset=unique(df$time_onset),
  reporting_piece_index=c(rep(1, 50), rep(2, 50))
)
snapshot_with_Rt_index_df <- snapshot_with_Rt_index_df %>%
  select(-cases_true) %>% 
  left_join(reporting_pieces, by="time_onset")
initial_Rt <- tribble(~Rt_index, ~Rt,
                      1, 1.5,
                      2, 1.5,
                      3, 0.4,
                      4, 1.5,
                      5, 1.5)
Rt_indices <- unlist(map(seq(1, 5, 1), ~rep(., 20)))

Rt_index_lookup <- tibble(
  time_onset=seq_along(Rt_indices),
  Rt_index=Rt_indices)
snapshot_with_Rt_index_df <- snapshot_with_Rt_index_df %>% 
  left_join(Rt_index_lookup)

initial_reporting_parameters <- tibble(
  reporting_piece_index=unique(reporting_pieces$reporting_piece_index)
) %>% 
  mutate(
    mean=c(5, 5),
    sd=3
  )
serial_parameters <- list(mean=5, sd=3)
Rt_prior <- list(shape=5, rate=5)
priors <- list(Rt=Rt_prior,
               reporting=list(mean_mu=5,
                              mean_sigma=10,
                              sd_mu=3,
                              sd_sigma=5),
               max_cases=10000)

res <- mcmc(niterations=100,
            snapshot_with_Rt_index_df,
            priors,
            serial_parameters,
            initial_cases_true,
            initial_reporting_parameters,
            initial_Rt,
            reporting_metropolis_parameters=list(mean_step=0.25, sd_step=0.1),
            serial_max=40, p_gamma_cutoff=0.99, maximise=FALSE,
            nchains = 1, is_parallel = FALSE)
```

The reporting delay means converge towards their true means.
```{r}
res$reporting %>% 
  group_by(reporting_piece_index) %>% 
  filter(iteration >= 50) %>% 
  summarise(mean=median(mean),
            sd=median(sd))

res$reporting %>% 
  ggplot(aes(x=iteration, y=mean)) +
  geom_line() +
  geom_hline(data=tibble(means=c(mean_1, mean_2),
                         reporting_piece_index=c(1, 2)),
                         aes(yintercept=means),
             linetype=2) +
  facet_wrap(~reporting_piece_index)
```


